---
title: "Run animal models in asreml"
author: "Carys Jones"
date: '2023-01-31'
---


# Dependencies
```{r}

box::use(.. / R / dirs[dirs])
box::use(mods = ../ R / for_models)

box::use(asreml)
box::use(dplyr)

```


# BREEDING DATA --------------------------------------------------------------


```{r}

gtit_data_sub_nb <- read.csv(file.path(dirs$data, 'GTIT_data_anim_mod.csv'), na.strings=c("", "NA"))
nrow(gtit_data_sub_nb) #11526

#make factors
gtit_data_sub_nb <- gtit_data_sub_nb %>%
   dplyr::mutate_at(c('breeding_year', 'Mother', 'Fbreeding_age_group', 'nest.box'), as.factor) 

#to run with a subset - if you want to mess around a run models with subset of years etc
#helpful when running big models to test things - I wrote a basic function to do that quickly
#put dates as range e.g. 2011:2021, or 'ALL' to get all years of data 
breed_data <- mod$choose_data(gtit_data_sub_nb, years = 'ALL')
#checks
unique(breed_data$breeding_year)
nrow(breed_data) #11526


```



# PEDIGREE DATA --------------------------------------------------------------

```{r}

#pedigree data input 
ped_data <- read.csv(file.path(dirs$data, 'GTIT-pedigree-for-asreml.csv'), na.strings=c("", "NA"))
nrow(ped_data) #15745

#get inverse of matrix - need this to run models
ped_data_inv <- asreml::ainverse(ped_data)


```


# RUN MODELS --------------------------------------------------------------

```{r}

#For laying date - running some basic models, shouldn't take too long to run 

#model basic  - 
#fixed effect of breeding age (as juvenile = 1, and adult = >1)
#random effect of breeding year, vm = additive genetic effect, ide = permanenet environment effect
LD_basic <- asreml::asreml(fixed = April.lay.date ~ 1 + Fbreeding_age_group,
                  random = ~vm(Mother, ped_data_inv) + 
                            ide(Mother, ped_data_inv) +
                            breeding_year,
                  data = breed_data)
#save output
saveRDS(LD_basic, file = file.path(dirs$output,'LD_basic.rds'))


#model with NB -
#same as above but now including the nest box ID as a random effect too 
LD_NB <- asreml::asreml(fixed = April.lay.date ~ 1 + Fbreeding_age_group,
                random = ~vm(Mother, ped_data_inv) + 
                  ide(Mother, ped_data_inv) +
                  breeding_year + 
                  nest.box,
                data = breed_data)
saveRDS(LD_NB, file = file.path(dirs$output,'LD_NB.rds'))

